---
title: |
  | Alpha diversity -- <br> Salmonella project metabarcoding data
pagetitle: "Alpha diversity"
author: "Jelmer Poelstra (poelstra.1@osu.edu), MCIC Wooster"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    highlight: kate
    toc: true
    toc_float: true
    fig_caption: true
    anchor_sections: true
    df_print: kable
    css: html_page.css
bibliography: salmonella.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  out.width = "90%"
)

knitr::opts_knit$set(root.dir = here::here(),
                     output.dir = here::here("reports/html"))
```

```{r}
## Load packages
if (!"pacman" %in% installed.packages()) install.packages("pacman")
packages <- c("remotes", "here",
              "tidyverse",
              "patchwork", "ggpubr", "RColorBrewer",
              "DT", "kableExtra", "modelsummary", "gt",
              "phyloseq", "microbiome", "ampvis2",
              "breakaway", "DivNet")
pacman::p_load(char = packages)
## remotes::install_github("MadsAlbertsen/ampvis2")
## remotes::install_github("adw96/breakaway")
## remotes::install_github("adw96/DivNet")
```

```{r}
## Source script with functions
source(here("scripts/report_funs.R"))
```

```{r}
## DATASET-SPECIFIC SETTINGS
trt_lev_all <- c("NC", "L", "S", "L+S")
trt_col_all <- brewer.pal(length(trt_lev_all), "Set1")
names(trt_col_all) <- trt_lev_all

trt_lev_noNC <- c("L", "S", "L+S")
trt_col_noNC <- brewer.pal(4, "Set1")[2:4]
names(trt_col_noNC) <- trt_lev_noNC

## Before rarefaction, remove the sample with the least nr of reads?
remove_least_sample <- FALSE

## Diversity measures
div_measures <- c("Observed", "Shannon", "Simpson")
```

```{r}
## General settings
theme_set(theme_bw(base_size = 14))
seed <- 100
set.seed(seed)
```

```{r}
## Define input files
ps_file <- here("results/phyloseq/ps_dadatax_filt.rds")
divnet_res_file <- here("results/divnet/divnet_by_ind.rds")
```

```{r}
## Load input files
ps <- readRDS(ps_file)
divnet_res <- readRDS(divnet_res_file)
```


```{r}
## Treatment factor levels
sample_data(ps)$treatment <- factor(
  sample_data(ps)$treatment, levels = trt_lev_all
  )

## Extract metadata
meta <- meta(ps) %>%
  mutate(sampleID2 = gsub("-", ".", sampleID))  # estimate_richness() will return these names
```

```{r}
# Phyloseq counts transformations

## Proportional transformation
ps_prop <- transform(ps, "compositional")

## Remove "orphan samples" -- only one for a given factor
fct_counts <- table(sample_data(ps)$treatment)
if (any(fct_counts == 1)) {
  fct_remove <- names(which(fct_counts == 1))
  ps_for_models <- subset_samples(ps, treatment != fct_remove)
  sample_data(ps_for_models)$treatment <- factor(
    sample_data(ps_for_models)$treatment, levels = trt_lev_noNC
    )
} else {
  ps_for_models <- ps
}

## Rarefaction
if (remove_least_sample == TRUE) {
  idx_least <- sample_sums(ps_for_models) == min(sample_sums(ps_for_models))
  smp_least <- names(sample_sums(ps_for_models)[idx_least])
  ps_for_rar <- subset_samples(ps_for_models, sampleID != smp_least)
} else {
  ps_for_rar <- ps_for_models
}

ps_rar <- rarefy_even_depth(ps_for_rar,
                            sample.size = min(sample_sums(ps_for_rar)),
                            rngseed = seed, replace = FALSE)
```

<br>

-----

## Methods

### Summary for manuscript

Taxonomic richness and diversity were first calculated using the `phyloseq`
`estimate_richness()` function,
and differences between treatments were tested for using a Kruskal-Wallis test.
Since these estimates do not correct for biases due to differences in
sequencing depth (the total number of reads) among samples,
this was done not only on the raw data but also on a rarefied dataset
(rarefaction was done with the `phyloseq` `rarefy_even_depth()` function).

Next, we used methods that directly try to account for differences in
sequencing depth when estimating richness and diversity.
Specifically, taxonomic richness was estimated for each sample and compared
between treatments using the R package `breakaway`
(version 4.7.6, [@willis_estimating_2015; @willis_improved_2017]),
while alpha-diversity (using the Simpson and Shannon indices)
was estimated and compared among treatments using the R package `DivNet`
(version 0.4.0, @willis_estimating_2020).

<br>

-----

## Richness v. sequencing depth {.tabset .tabset-fade .tabset-pills}

### Rarefaction plot

**A good sign is that these plots flatten out for most samples --**
**but not for the two samples with just a few hundred reads.**

```{r}
metadata <- data.frame(sample_data(ps), check.names = FALSE)
asv_table <- data.frame(t(otu_table(ps)), tax_table(ps),
                        check.names = FALSE)
ps_amp <- amp_load(asv_table, metadata)
```

```{r}
amp_rarecurve(ps_amp, stepsize = 10, color_by = "treatment") +
  labs(x = "Total ASV count (library size)",
       y = "Number of distinct ASVs (richness)",
       color = "treatment") +
  scale_color_manual(values = trt_col_all) +
  scale_x_continuous(expand = c(0, 0), labels = scales::comma) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)))
```

<br>

----

### Correlation plot

**There is a significant correlation between sequencing depth and richness.**
(But this goes away when the two lowest-count samples are removed -- not shown.)

```{r}
## Check whether sequence depth and total richness are correlated across samples
divs <- data.frame(total_count = sample_sums(ps),
                   n_ASV = rowSums(otu_table(ps) != 0)) %>%
  rownames_to_column("sampleID") %>%
  left_join(meta(ps), by = "sampleID")

ggplot(divs) +
  aes(x = total_count, y = n_ASV) +
  ggpubr::stat_cor(method = "spearman", size = 4,
                   label.x.npc = 0.2, label.y.npc = 0.99) +
  geom_point(aes(color = treatment)) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", size = 0.5) +
  scale_color_manual(values = trt_col_all) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Total ASV count (sequence depth)",
       y = "Number of distinct ASVs (richness)")
```

<br>

----

## Alpha diversity with _phyloseq_

"Observed" diversity = richness = the number of distinct ASVs,
not taking evenness into account.

**No significant differences between treatments were found.**

### Rarefied data {.tabset .tabset-fade .tabset-pills}

Used data rarefied to the minimum number of reads for any one sample.

**No significant differences between treatments were found.**

```{r}
rich_df <- estimate_richness(ps_rar, measures = div_measures) %>%
  merge(., meta, by.x = "row.names", by.y = "sampleID2")
```

#### Plot

```{r}
plot_richness(ps_rar,
              x = "treatment", color = "treatment",
              measures = div_measures) +
  geom_point(size = 2.5) +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
```

#### Model: richness (observed) 

```{r}
kruskal.test(Observed ~ treatment, data = rich_df)
```

#### Model: Shannon

```{r}
kruskal.test(Shannon ~ treatment, data = rich_df)
```

#### Model: Simpson

```{r}
kruskal.test(Simpson ~ treatment, data = rich_df)
```

### Raw data {.tabset .tabset-fade .tabset-pills}

```{r}
rich_df <- estimate_richness(ps, measures = div_measures) %>%
  merge(., meta, by.x = "row.names", by.y = "sampleID2")
```

#### Plot

```{r}
plot_richness(ps,
              x = "treatment", color = "treatment",
              measures = div_measures) +
  geom_point(size = 2.5) +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
```

#### Model: richness (observed) 

```{r}
kruskal.test(Observed ~ treatment, data = rich_df)
```

#### Model: Shannon

```{r}
kruskal.test(Shannon ~ treatment, data = rich_df)
```

#### Model: Simpson

```{r}
kruskal.test(Simpson ~ treatment, data = rich_df)
```

<br>

-----

## Richness with _breakaway_ {.tabset .tabset-fade .tabset-pills}

Here, we compare species _richness_ (i.e. just the number of taxa)
estimated with the R package _breakaway_
([paper](https://onlinelibrary.wiley.com/doi/abs/10.1111/biom.12332),
[package](https://github.com/adw96/breakaway))
among treatments.

**According to this test, richness is significantly lower for the NC and L treatments**
**compared to L+S.**
**NC also has lower richness than S.**

### Boxplot

```{r}
plot_div(ps_for_models, covar = "treatment") +
  scale_color_manual(values = trt_col_all)
```

### Plot with per-sample error bars

Note that richness estimation is done on a _per-sample_ basis,
so we can have error bars for each sample:

```{r}
plot_div(ps_for_models, each_sample = TRUE) +
  scale_color_manual(values = trt_col_all)
```

### Model -- L+S vs others

```{r}
## All 3 treatment levels
rich_res <- get_break(ps_for_models) %>% 
  mutate(treatment = relevel(factor(treatment), ref = c("L+S")))

bt_treat <- betta(formula = estimate ~ treatment, ses = error, data = rich_res)

modtab(bt_treat, model_names = "All treatments")
```

### Model -- NC vs others

```{r}
rich_res <- get_break(ps_for_models) %>% 
  mutate(treatment = relevel(factor(treatment), ref = c("NC")))

bt_treat <- betta(formula = estimate ~ treatment, ses = error, data = rich_res)

modtab(bt_treat, model_names = "All treatments")
```

<br>

-----

## Alpha diversity with _DivNet_ {.tabset .tabset-fade .tabset-pills}

**According to this test, Simpson diversity is significantly lower for the**
**S treatment compared to L+S and to N+C, among others.**

### Boxplot

```{r}
p_sh <- plot_div(ps_for_models, divnet_res, div_index = "shannon", covar = "treatment") +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
p_si <- plot_div(ps_for_models, divnet_res, div_index = "simpson", covar = "treatment") +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
p_sh + p_si
```

### Plot with per-sample error bars

Note that richness estimation is done on a _per-sample_ basis,
so we can have error bars for each sample:

```{r}
p_sh <- plot_div(ps_for_models, divnet_res,
                 div_index = "shannon", covar = "treatment",
                 each_sample = TRUE) +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
p_si <- plot_div(ps_for_models, divnet_res,
                 div_index = "simpson", covar = "treatment",
                 each_sample = TRUE) +
  scale_color_manual(values = trt_col_all) +
  guides(color = "none")
p_sh + p_si
```

### Model -- L+S vs others

```{r}
## Shannon
shan <- data.frame(estimate = summary(divnet_res$shannon)$estimate,
                   error = sqrt(divnet_res$`shannon-variance`)) %>%
  merge(meta(ps_for_models), ., by = "row.names") %>%
  mutate(treatment = relevel(factor(treatment), ref = c("L+S")))
shan_mod <- betta(formula = estimate ~ treatment, ses = error, data = shan)

## Simpson
simp <- data.frame(estimate = summary(divnet_res$simpson)$estimate,
                   error = sqrt(divnet_res$`simpson-variance`)) %>%
  merge(meta(ps_for_models), ., by = "row.names") %>%
  mutate(treatment = relevel(factor(treatment), ref = c("L+S")))
simp_mod <- betta(formula = estimate ~ treatment, ses = error, data = simp)

## Print formatted table
modtab(betta_res = list(shan_mod, simp_mod),
       model_names = c("Shannon", "Simpson"),
       fmt = 3)
```

### Model -- NC vs others

```{r}
## Shannon
shan <- data.frame(estimate = summary(divnet_res$shannon)$estimate,
                   error = sqrt(divnet_res$`shannon-variance`)) %>%
  merge(meta(ps_for_models), ., by = "row.names") %>%
  mutate(treatment = relevel(factor(treatment), ref = c("NC")))
shan_mod <- betta(formula = estimate ~ treatment, ses = error, data = shan)

## Simpson
simp <- data.frame(estimate = summary(divnet_res$simpson)$estimate,
                   error = sqrt(divnet_res$`simpson-variance`)) %>%
  merge(meta(ps_for_models), ., by = "row.names") %>%
  mutate(treatment = relevel(factor(treatment), ref = c("NC")))
simp_mod <- betta(formula = estimate ~ treatment, ses = error, data = simp)

## Print formatted table
modtab(betta_res = list(shan_mod, simp_mod),
       model_names = c("Shannon", "Simpson"),
       fmt = 3)
```

<br>

----

## References
